Cartogram Mapping and its Application to Cancer Data Visualization

### 2019-09-30 review_article.Rmd
cartograms done

\section*{Abstract}
Choropleth maps are widely used in cancer atlases, and traditionally communicate cancer statistics over geographic domains. The type of statistic displayed varies immensely, and effective choices can assist in the understanding of impact of cancer on a population. Atlases have moved from print media to web format, and the availability of graphical user interfaces, and user interaction and animation, greatly enhance the ability to communicate more information and positively engage readers. More use of alternatives to choropleth maps, such as cartograms, could be employed in online atlases, for better representation of spatial distributions, and show statistics on small areas. This is especially true for the geography of Australia, where the population is concentrated is small areas along the coastlines.

# Introduction
<!-- What data are we working with, why, what are the purposes -->
Cancer statistics are usually delivered as an aggregated value for a geopolitical area. Presenting these statistics requires transforming individual observations into aggregations of communities as geographical units, in large part for privacy protection or political and policy purposes.  The information could be as simple as counts per area (e.g. state, province, local government area, post/zip code). Counts alone are not sufficient to compare areas, because the populations of areas are all different. In this case, the counts data needs to be merged with population data to appropriately calibrate it to incidence, for example, rate per 100000 people.  This type of data is collected on a routine basis for public health purposes, and may be made available to the general public as a service to the community. The task, then, is to examine what are the usual ways to communicate cancer statistics, to the public, are there alternative approaches, and what are the pros and cons of these choices.

<!-- Ways to visualize, whats our hypothesis -->
A common approach to communicate cancer statistics, is to display statistics on a map. Using a choropleth map: the statistic is mapped to color and the geographic region is filled with this color. The viewer would then be able to examine the spatial distribution of the disease incidence, where there is a trend in longitude or latitude, or rural vs urban, or coastal vs inland, or even specific hot spots of the disease. Visualizing diseases on maps is often the first step in exploratory spatial data analysis and effectively helps in the formulation of hypotheses [@MTMSIH]. Disease maps help to present geographic patterns that may be overlooked in a table, obscuring the geospatially related statistics [@SAMGIS].
By providing a visual representation of cancer outcomes, geographic patterns of disease may be identified and effectively addressed with public health policy and actions.
@SE recognizes one of the key challenges with mapping spatial patterns of disease is the design of visualizations.
This paper addresses the visualization techniques and their applications to cancer statistics. Highlighting the differences and historic use of these displays.

<!-- Structure of the paper -->
The paper is organized as follows. The next section describes the choropleth map which is the common approach to disease maps. Section \ref{ch:public} surveys atlases in use today. Section \ref{ch:alternatives} describes an alternative display, the cartogram what may be useful for countries that have heterogeneously sized geographic units. The pros and cons of these approaches is discussed in Section \ref{ch:compare}. Disease maps are more useful when made interactive, and common options are described in Section \ref{ch:interacting}, along with a discussion of benefits and disadvantages. The last section summarizes the paper and discusses future directions.

# Map displays for disease data {#ch:choropleths}

<!-- what is a choropleth map -->
A choropleth map is used to display differences in the geographical distribution of data by spatial unit by shading areas of a map. The geography is faithfully rendered, and the color rendering is designed to reveal spatial patterns among data values. A choropleth is constructed by drawing the geographic or political boundaries, and filling the shapes with colors to represent values of a measured variable [@EI]. 
Figure \ref{fig:choroCRS} shows a choropleth of age-adjusted rate (per 100,000 people) of new cases of lung and bronchus in the USA, averaged over 2012 through 2016. The data was extracted from the official federal statistics by @USInteractive on cancer incidence and deaths, produced by the Centers for Disease Control and Prevention (CDC) and the National Cancer Institute (NCI).

<!-- history and types of choropleths, variations -->
Early versions of choropleth maps used symbols or patterns instead of color.
@CPISACA discuss the use of choropleths to visualize cancer data, and @DMAHP gives an overview of the development of these maps for displaying disease data.

<!-- psychology -->
Utilizing the state boundaries can make a map familiar to read [@CIBMUK], and allows viewers to visually infer the spatial relationships in the data. 
The familiarity of the geography is a worthy consideration when presenting results of spatial analysis. Just as geographers are no longer the only creators of maps, @CPISACA suggests the audiences of spatial health data analysis have extended beyond researchers to the public, policymakers and the media.

<!-- how do we use choropleth's -->
Identifying and explaining spatial structures, patterns, and processes involves considering the individuals in communities and organizing communities into representable units [@SAMGIS]. In Figure \ref{fig:choroCRS}, a west to east spatial trend of increasing rates, can be seen. There is also a spatial outlier -- Utah has a noticeably lower rate that its neighbors. Also Kentucky has a noticeably high rate, and Maine also has a higher rate than its neighbors. There is something of a cluster of higher rates around the tobacco states.

<!-- potential problems -->
While the areas are recognizable shapes, they are often politically driven boundaries with individual areas being of non-uniform size, containing different population densities and subject to change over time. 
The different population and geographical sizes of administrative areas can attract attention to the shades of the underpopulated but large areas [@EI], Skowronnek [-@BCM] calls this an area-size bias. 
Choropleths can inhibit visual inference when presenting human related statistics as the display may draw attention away from the 'potentially more important results in the more populous communities' that are geographically smaller [@SE].

@@

# Cancer atlases {#ch:public}

<!-- Displaying of cancer data on choropleths -->
Choropleth maps can be useful devices for communicating information to public on a familiar map base.
A cancer atlas is a choropleth map, or collection of maps, representing cancer incidence and mortality for a country, or group of countries. 
In epidemiology, choropleths are often used as a tool to study the spatial distribution of cancer incidence and mortality. 
The data collection methods of cancer mortality rates across regions, and the administrative control within regions lends itself to choropleth visualization. 
@MACM provides the definition of a cancer atlas, beginning with Haviland's maps in 1875, they attribute UK cancer atlases to Howe (1963), and early work in US cancer atlases can be attributed to Burbank (1971).
The increasing development and use of disease maps can be attributed to the availability of geographic information system software [@SE].
The choropleth maps presented levels via hatching or dots on a black and white scale.
These atlases were key to developing hypotheses regarding areas with unusually high rates, geographic correlations, work related exposures, and high risk diets [@MACM].

<!-- Measures reported in cancer atlases -->
The presentation of cancer statistics has changed over time with greater access to computational power.
Mortality rates are now often presented as relative rates of risk across the population, and age adjusted to correct for the higher prevalence of cancers in older populations.
Howe [-@HEDP] describes Stock's development of the standardized mortality ratios through the 1930s. Table \ref{tab:measures} summarizes the measures presented in published cancer atlases, and provides a definition of each measure.

Table: \label{tab:measures} Common statistics for reporting cancer information.

|Measure |Details| 
|:--------------------|:------------------------------------------------|  
|1. Count|Crude cancer counts|
|2. Rate per 100,000|Cancer incidence per 100,000 population|  
|3. New cancer cases per 100,000|Specific methods could not be found|   
|4. IR (Incidence Ratio)|$(IR)_i=\frac{(Incidence\ Rate)_i}{Average\ Incidence\ Rate}$, |
||Cancer incidence rate in region $i$ over the average cancer incidence rate for the total region|
|5. Age Adjusted Rate per 100,000| Standardized by age structure or region|   
|6. Age Adjusted Relative Risk| RR standardized by age structure in each region $i$|  
|7. SIR (Standardized Incidence Ratio)| IR standardized by age structure in each region $i$|
|8. Below or above Expected|Alternative expression of the SIR|
|9. RER |$RER = \frac{(Cancer\ related\ mortality)_i}{Average\ cancer\ related\ mortality}$|
|(Relative Excess Risk)|Represents the estimate of cancer related mortality within five years of diagnosis|
||Also referred to as 'excess hazard ratio'|

@@

Cancer maps are effective tools for communicating to wide range of audiences, including the general public and others not trained in statistical analyses.
These visualizations enable non-expert audiences to interpret the outputs of sophisticated statistical analyses. Cruickshank (1947) as cited by @DMAHP, discusses using visuals as a 'formal statistical assessment of the spatial pattern'. Overwhelmingly, cancer maps utilized to communicate to the public and other non-expert audiences are choropleths.  These atlases provided Incidence, Survival and Mortality rates to the public on areal map bases.

The Atlas of Cancer in Queensland [@DACQ] focused on highlighting the difference in experience for those living in rural and disadvantaged areas, the Standardized Incidence Ratios (SIR) were modeled via Bayesian methods.
The presentation of these rates considers not only the shapes of the areas, but also appropriate choices of color blind friendly color schemes, and categories of the values to highlight significantly different areas.

## Overview of publicly available atlases \label{sec:publicatlases}

@roberts2019communication identified 33 publicly available cancer atlases, published between January 2010 and November 2015. All of these use choropleth maps. All but one of these were published by non-commercial organizations, including not-for-profits, government, research organizations, advocacy groups or government funded partnerships. The use of choropleths within the public domain mirrors the heavy use of choropleth maps within the research literature, discussed above. 
The cancer atlases identified by @roberts2019communication covered geographies from all around the world, most focused on single nations. Figure \ref{fig:choropleth-grid-create} displays a global (b), national (a, d, e, f) and state (c, g) choropleths. The sections below provide details on each of the maps within Figure \ref{fig:choropleth-grid-create}.

@@

Figure \ref{fig:choropleth-grid-create}a contains an image from *[The Environmental and Health Atlas of England and Wales](http://www.envhealthatlas.co.uk/eha/Breast/)* [@EnvEnglandWales2]. This map shows the relative risk for women developing lung cancer in England and Wales in 2010. The cancer data used to generate this map came from Office for National Statistics (ONS) (England) and from the Welsh Cancer Intelligence and Surveillance Unit (WCISU).

@@

The map seen in Figure \ref{fig:choropleth-grid-create}b is from the *[Globocan 2012: Estimated  Cancer Incidence, Mortality and Prevalence](http://globocan.iarc.fr/Pages/Map.aspx)* [@Globocan]. This global map shows age standardized incidence rates (per 100,000) for all invasive cancers for both men and women, aggregated at a national level for 2018. This map is published by the [World Health Organization's International Agency for Research on Cancer](https://www.iarc.fr/). Data was sourced from cancer registries in each country, contributing registries can be seen in the supplementary material on the cancer atlas website.

@@

Figure \ref{fig:choropleth-grid-create}c shows an extract from  *[The Atlas of Cancer in QLD](https://cancerqld.org.au/research/queensland-cancer-statistics/queensland-cancer-atlas/)* [@QLDcancerAtlas]. This map was published by the Queensland Cancer Council and shows the relative incidence ratio of lung cancer in males in the state of QLD within Australia based on data from 1998 to 2007. Data to generate this map was sourced from the *[Queensland Cancer Registry](https://cancerqld.org.au/research/cancer-registries/queensland-cancer-registry/)*.

### Bowel Cancer Australia Atlas 
Figure \ref{fig:choropleth-grid-create}d shows the *[Bowel Cancer Australia Atlas](http://www.bowelcanceratlas.org/)* [@Bowel]. Published by *[Bowel Cancer Australia](https://www.bowelcanceraustralia.org/)* (Australia). This map shows the percentage of Australian males between 50 - 54 years of age that diagnosed with bowel cancer in 2016 in Australia. The source of the data is not provided.

### United States Cancer Statistics: An Interactive Cancer Statistics Website
The *[United States Cancer Statistics: An Interactive Cancer Statistics Website](https://nccd.cdc.gov/DCPC_INCA/)* [@USInteractive] can be seen in Figure \ref{fig:choropleth-grid-create}e. This map contains  the incidence rate per 100,000, of all cancer types for men and women in the United States in 2016, aggregated at the state level. The map was published by the *[Centers for Disease Control and Prevention](https://www.cdc.gov/)*. Incidence data seen in this map were compiled from cancer registries meeting U.S. Cancer Statistics data quality criteria covering 100% of the U.S. population.

### Map of Cancer Mortality Rates in Spain
The *[Map of Cancer Mortality Rates in Spain](http://elpais.com/elpais/2014/10/06/media/1412612722_141933.html)* [@cancerSpain] can be seen in Figure \ref{fig:choropleth-grid-create}f. The side by side maps show relative risk of lung cancer for men vs women based on data from 2004 to 2008. The source of the data and statistical methods are unknown.

### Atlas of Childhood Cancer in Ontario 
Figure \ref{fig:choropleth-grid-create}g shows an extract from the Atlas of Childhood Cancer in Ontario and specifically displays the *[incidence rate of childhood cancers](https://www.pogo.ca/wp-content/uploads/2015/02/POGO_CC-Atlas-3-Incidence_Feb-2015.pdf)* [@OntarioPediatric] per 100,000 (by census division) for children aged 0-14, in Ontario from 1995 to 2004.

## Common statistics displayed

Cancer maps are powerful visualizations that summarize complex statistical analyses, however the statistics represented in these maps cannot tell the entire story.
Supplementary graphs and plots are often included to add more depth and information to the map.
@CPISACA suggests additional materials such as tables, graphs, and text explanations support understanding and inference derived from maps, ensuring the message communicated will be consistent across a range of viewers.
There are many visualizations used for displays of statistical summaries, these may be dot plots, bar plots, box plots, times series plots, cumulative distribution plots, scatter plots, Q-Q plots. 
These additional displays of the cancer distribution can provide alternative views of the cancer statistics, as well as the supporting statistics including error, confidence intervals, distributions, sample or population sizes, standard deviation and other measures.
When presenting cancer maps, @MACM believes the intuition derived from maps must be 'validated by rigorous statistical analyses', the supplementary statistics help for this validation.

@@

It is extremely likely that each population area will have a different number of people. The distribution of the population residing in all areas may also be communicated in a table or histogram display [@NICR_1].
Atlases can connect the population to the land available to them by communicating population density.
Atlases can also connect the population to the land available to them by communicating population density.

### Statistical uncertainty
Additional statistics that accompany an atlas often include a measure of the statistical uncertainty surrounding the statistics presented in a choropleth.
In the review of atlases in the public domain, atlases were considered to report uncertainty to the non-expert user if they included a measure of statistical uncertainty either within or alongside the map [@roberts2019communication]. 
The maps considered used standard and well-known measures including credible intervals and standard deviation, statistical significance, box plots and distributions. Other methods involve providing adjacent maps or overlapping maps with symbols [@VSSDCUC].
The maps employing uncertainty ranged from static documents or infographics, to interactive online visualizations. Communicating the statistical uncertainty associated with the estimates occurs using confidence intervals (CI), credible intervals (CrI), statistical significance levels, box plots, distribution plots, and reporting sample size and standard deviations.
Close to half of the atlases identified (42%, n=14) included some measure of uncertainty. The most common measure used to represent uncertainty were credible or confidence intervals (CIs).

### Demographics
Demographics include information regarding the age and sex distribution of the areas presented. 
<!-- sex -->
Sex is an important cofactor for cancer atlases. 
As some cancers are sex specific, and others may be found in both males and females, atlases often specify the relevant sex as part of the visual output in the displays. 
Digital atlases allow for users to interact with the controls of the displays, they can select males, females or both depending on the type of cancers explored.

### Socio-economic indicators
Socio-economic indicators can explain how the experience of cancer prevalence varies for various members of a society.
These indicators include unemployment rates, poverty rates, remoteness, and education levels achieved though, only a few atlases also explored the impact of rurality on cancer rates. 
These rates may also be explored as percentages above or below the mean or median value for the set of spatial areas.
The Human Development Index can be used to understand the socio-economic experience of a community, as well as Income levels which can be sourced from the World Bank list of economies [@IARC_3]. 
The areas are often ranked and allocated to quintiles; each quintile can be presented as categories describing the ranking.

One of the concerns of adding too much information to a map is the fear of cognitive overload [@mcgranaghan1993cartographic] in which the user reaches an information threshold, beyond which will not be able to make sense of the information. These concerns are not unfounded, complexity and density of representation methods appear to overwhelm novice decision makers, while experts are able to use the detail more readily when making-decisions [@cliburn2002design]. Interactivity is a design feature within modern mapping methods that can be used to incorporate additional information and complexity without overloading the user.

Interactivity enables supplementary information to be incorporated into online atlases without cluttering the display. Interactive design features found in online cancer maps include tool tip features, drop down menus, data selection, zooming and panning allow users to explore the map as they want more information and allow flexibility in the display [@HTLWM]. The use of these supports were found in various online cancer maps identified by @roberts2019communication. The controls for basic interactive features are often placed outside of the plot space [@TGA], thus the map image is updated/replaced as the user interacts with the controls. For example, changing the population age or other demographic variables. Some more advanced interactions include direct interactions with the plot via the use of overlaid tool tip features, very few cancer atlases involve these more complex selection tools.

These interactive features provide an opportunity for users to explore the additional information available. This helps users to understand and interpret the spatial distribution presented, as well as validate, explain or explore the presented statistics and their relationships to each other and/or their underlying spatial distribution. This allows relationships between spatial areas and diseases to be explored with sophistication in non-traditional but still 'cognitively accessible' ways [@TNTEA]. The interactive features of the publicly available maps identified by @roberts2019communication allow the exploration of geographic hierarchies, population distribution, statistical uncertainty, demographics and socio-economic indicators.
@TNTEA suggested LM plots as a solution to linking cartography and statistical graphics.

@@

## Cartograms

<!-- What is a cartogram? -->
Choropleths imply uniformity of data across the geographic space however population densities are extremely unlikely to be uniform [@BCM].
A cartogram displays abstracted geographic space, with the intention of improving the presentation of the statistic of interest. For a single variable of interest, each map area is changed to emphasize the distribution by representing the corresponding value, in comparison to the value of the other areas [@ACCAC].
Changes in the map base are implemented by altering the boundaries, and therefore shapes, of individual areas.

<!-- There are many alternative views of the world-->
<!-- Tobler map projections-->
Choropleths may be considered true topological maps, however, if the land mass displayed covers enough of the globe, there must be a transformation or distortion to display the land in 2D [@HTLWM]. The amount of distortion is related to the distance covered by the landmass displayed @GAMP.
World map projections reflect the frequent perspectives used to view the earth. Choropleth maps will always be distorted if they cover enough of the globe, just like photographs of the globe from space. Choropleth creation requires choosing a map projection that shows a favorable distortion of the geography for presenting the set of spatial information. Selecting a display can prevent misinterpretation of global statistics, as global maps face the challenge of equitable displays of land mass on maps [@RSCW].
If the statistic presented on the map base relies on physical distance and is influenced by the topology there is no transformation needed, beyond choosing a reasonable projection.

```{r choroCRS, fig.cap = "Four choropleth maps of the United States of America using various coordinate reference systems. Each state has been colored according to the average age-adjusted rate of incidence for lung and bronchus for females and males in the United States 2012-2016. The map projections alter the shapes and angles of the boundaries of each state. Maps a and b are similar in their straight edges, unlike maps c and d which curve on the northern United States border.", message=FALSE, warning=FALSE, out.width='100%', fig.align='center'}
#knitr::include_graphics("figures/ggchoroCRS.png")
c1 <- png::readPNG("figures/gg3857.png")
c2 <- png::readPNG("figures/gg2163.png")
c3 <- png::readPNG("figures/gg4326.png")
c4 <- png::readPNG("figures/gg2955.png")
#knitr::include_graphics("figures/usa_grid.png")
load("figures/usa_legend.rda")

@@

<!-- Denominators -->
Event cartograms change the area of regions on a map depending on the amount of disease related events, but this does not consider the effects of land area and population [@VSSDCUC].
The purposeful distortion of the map space, transformed according to population density, is beneficial when a uniform density of the map base is desired. Population then becomes a uniformly distributed background for the statistic presented [@GOINO]. @ACTUC suggests 'population distribution is often extremely uneven in former British colonies', this makes the distortion necessary [@CTTMB].
When implementing a distortion of the geographical shape according to population, the resulting display is an area cartogram [@NAC], or population-by-area cartogram [@TAAM].

>>>>

\section*{Abstract}
<!-- 119 words -->
Choropleth maps are widely used in cancer atlases, and traditionally communicate cancer statistics over geographic domains. The type of statistic displayed varies immensely, and effective choices can assist in the understanding of impact of cancer on a population. Atlases have moved from print media to web format, and the availability of graphical user interfaces, and user interaction and animation, greatly enhance the ability to communicate more information and positively engage readers. More use of alternatives to choropleth maps, such as cartograms, could be employed in online atlases, for better representation of spatial distributions, and show statistics on small areas. This is especially true for the geography of Australia, where the population is concentrated is small areas along the coastlines.

# Introduction
<!-- 436 words -->
<!-- What data are we working with, why, what are the purposes -->
Cancer statistics are usually delivered as an aggregated value for a geopolitical area. Presenting these statistics requires transforming individual observations into aggregations of communities as geographical units, in large part for privacy protection or political and policy purposes.  The information could be as simple as counts per area (e.g. state, province, local government area, post/zip code). Counts alone are not sufficient to compare areas, because the populations of areas are all different. In this case, the counts data needs to be merged with population data to appropriately calibrate it to incidence, for example, rate per 100000 people.  This type of data is collected on a routine basis for public health purposes, and may be made available to the general public as a service to the community. The task, then, is to examine what are the usual ways to communicate cancer statistics, to the public, are there alternative approaches, and what are the pros and cons of these choices.

<!-- Ways to visualize, whats our hypothesis -->
A common approach to communicate cancer statistics, is to display statistics on a map. Using a choropleth map: the statistic is mapped to color and the geographic region is filled with this color. The viewer would then be able to examine the spatial distribution of the disease incidence, where there is a trend in longitude or latitude, or rural vs urban, or coastal vs inland, or even specific hot spots of the disease. Visualizing diseases on maps is often the first step in exploratory spatial data analysis and effectively helps in the formulation of hypotheses. Disease maps help to present geographic patterns that may be overlooked in a table, obscuring the geospatially related statistics [@SAMGIS].
By providing a visual representation of cancer outcomes, geographic patterns of disease may be identified and effectively addressed with public health policy and actions.
@SE recognizes one of the key challenges with mapping spatial patterns of disease is the design of visualizations.
This paper addresses the visualization techniques and their applications to cancer statistics. Highlighting the differences and historic use of these displays.

<!-- Structure of the paper -->
The paper is organized as follows. The next section describes the choropleth map which is the common approach to disease maps. Section \ref{ch:public} surveys atlases in use today. Section \ref{ch:alternatives} describes  alternative displays, including the cartogram which is useful when the map has heterogeneously sized geographic units. The pros and cons of these approaches is discussed in Section \ref{ch:compare}. Disease maps are more useful when made interactive, and common options are described in Section \ref{ch:interacting}, along with a discussion of benefits and disadvantages. The last section summarizes the paper.

# Map displays for disease data {#ch:choropleths}
<!-- 381 words -->
<!-- what is a choropleth map -->
A choropleth map is used to display differences in the geographical distribution of data by spatial unit by shading areas of a map [@EI]. The geography is faithfully rendered, and the color rendering is designed to reveal spatial patterns among data values. Identifying and explaining spatial structures, patterns, and processes involves considering the individuals in communities and organizing communities into representable units [@SAMGIS].
Early versions of choropleth maps used symbols or patterns instead of color.
@CPISACA discuss the use of choropleths to visualize cancer data, and @DMAHP gives an overview of the development of these maps for displaying disease data. In epidemiology, choropleths are often used as a tool to study the spatial distribution of cancer incidence and mortality.

Figure \ref{fig:choroCRS} shows various geographic projections of choropleth maps of age-adjusted rate (per 100,000 people) of new cases of lung and bronchus in the USA, averaged over 2012 through 2016. The data was extracted from the official federal statistics by @USInteractive on cancer incidence and deaths, produced by the Centers for Disease Control and Prevention (CDC) and the National Cancer Institute (NCI). A west to east spatial trend of increasing rates, can be seen. There is also a spatial outlier -- Utah has a noticeably lower rate that its neighbors. Also Kentucky has a noticeably high rate, and Maine also has a higher rate than its neighbors. There is something of a cluster of higher rates around the tobacco states.

<!-- psychology -->
Utilizing the state boundaries can make a map familiar to read [@CIBMUK], and allows viewers to visually infer the spatial relationships in the data. 
The familiarity of the geography is a worthy consideration when presenting results of spatial analysis, to consumers that extend beyond researchers, to the public, policymakers and the media [@CPISACA].

<!-- potential problems -->
While the areas are recognizable shapes, they are often politically driven boundaries with individual areas being of non-uniform size, containing different population densities and subject to change over time. 
The different population and geographical sizes of administrative areas can attract attention to the shades of the underpopulated but large areas [@EI], Skowronnek [-@BCM] calls this an area-size bias. 
Choropleths can inhibit visual inference when presenting human-related statistics as the display may draw attention away from the 'potentially more important results in the more populous communities' that are geographically smaller [@SE].

@@

# Cancer atlases {#ch:public}
<!-- 360 words -->
<!-- Displaying of cancer data on choropleths -->
A cancer atlas is a map, or collection of maps, representing cancer incidence and mortality for a country, or group of countries. Atlases are key to developing hypotheses regarding areas with unusually high rates, geographic correlations, work related exposures, and high risk diets [@MACM]. The data collection of statistics across regions, and the administrative control within regions, lends itself to choropleth visualization. 
Atlases date to Haviland's maps in 1875, with UK cancer atlases appearing in 1961 [@HEDP], and early work in US cancer atlases appearing in 1971 [@burbank]. The presentation of cancer statistics has increased with greater access to computational power, and the availability of geographic information systems software [@SE].

<!-- Measures reported in cancer atlases -->
Mortality rates are now often presented as relative rates of risk across the population, and age adjusted to correct for the higher prevalence of cancers in older populations.
Howe [-@HEDP] describes Stock's development of the standardized mortality ratios through the 1930s. Table \ref{tab:measures} summarizes the measures commonly presented in published cancer atlases.

Table: \label{tab:measures} Commonly used measures for reporting cancer information.

|Measure |Details| 
|:--------------------|:------------------------------------------------|  
|1. Count|Crude cancer counts|
|2. Rate per 100,000|Cancer incidence per 100,000 population|  
|3. IR (Incidence Ratio)|$(IR)_i=\frac{(Incidence\ Rate)_i}{Average\ Incidence\ Rate}$, |
||Cancer incidence rate in region $i$ over the average cancer incidence rate for the total region|
|4. Age Adjusted Rate per 100,000| Standardized by age structure or region|   
|5. Age Adjusted Relative Risk| Standardized by age structure in each region $i$|  
|6. SIR (Standardized Incidence Ratio)| Standardized by incidence in each region $i$|
|7. Below or above Expected|Alternative expression of the SIR|
|8. RER |$RER = \frac{(Cancer\ related\ mortality)_i}{Average\ cancer\ related\ mortality}$|
|(Relative Excess Risk)|Represents the estimate of cancer related mortality within five years of diagnosis. Also referred to as 'excess hazard ratio'|

@@

Cancer maps are effective tools for communicating incidence, survival and mortality to wide range of audiences, including the general public and others not trained in statistical analyses.
These visualizations enable non-expert audiences to interpret the outputs of sophisticated statistical analyses. Cruickshank (1947) as cited by @DMAHP, discusses using visuals as a 'formal statistical assessment of the spatial pattern'. Overwhelmingly, the cancer maps utilized to communicate to the public and other non-expert audiences are choropleths.

<!-- written below The Atlas of Cancer in Queensland [@DACQ] focused on highlighting the difference in experience for those living in rural and disadvantaged areas, the Standardized Incidence Ratios (SIR) were modeled via Bayesian methods.
The presentation of these rates considers not only the shapes of the areas, but also appropriate choices of color blind friendly color schemes, and categories of the values to highlight significantly different areas. 
-->

## Overview of publicly available atlases \label{sec:publicatlases}
<!-- 416 words -->
Roberts [@roberts2019communication] identified 33 publicly available cancer atlases, published between January 2010 and November 2015, from different countries around the globe. All of these use choropleth maps. All but one of these were published by non-commercial organizations, including not-for-profits, government, research organizations, advocacy groups or government funded partnerships.  Figure \ref{fig:choropleth-grid-create} displays a subset of maps from these atlases.

@@

Figure \ref{fig:choropleth-grid-create}a shows the relative risk for women developing lung cancer in England and Wales in 2010 [@EnvEnglandWales2]. The cancer data was from Office for National Statistics (ONS) (England) and from the Welsh Cancer Intelligence and Surveillance Unit (WCISU).

@@

The map seen in Figure \ref{fig:choropleth-grid-create}b shows age standardized incidence rates (per 100,000) for all invasive cancers for both men and women, aggregated at a national level for 2018 [@Globocan]. It is published by the [World Health Organization's International Agency for Research on Cancer](https://www.iarc.fr/). Data was sourced from cancer registries of each country.

@@

Figure \ref{fig:choropleth-grid-create}c shows the relative incidence ratio of lung cancer in males in the state of QLD within Australia based on data from 1998 to 2007 [@QLDcancerAtlas]. It is published by the Queensland Cancer Council, using data from the *[Queensland Cancer Registry](https://cancerqld.org.au/research/cancer-registries/queensland-cancer-registry/)*.

### Bowel Cancer Australia Atlas 
Figure \ref{fig:choropleth-grid-create}d shows the percentage of Australian males between 50 - 54 years of age diagnosed with bowel cancer in 2016 in Australia [@Bowel]. It is published by *[Bowel Cancer Australia](https://www.bowelcanceraustralia.org/)*.

### United States Cancer Statistics: An Interactive Cancer Statistics Website

The *[United States Cancer Statistics: An Interactive Cancer Statistics Website](https://nccd.cdc.gov/DCPC_INCA/)* [@USInteractive] can be seen in Figure \ref{fig:choropleth-grid-create}e. This map contains  the incidence rate per 100,000, of all cancer types for men and women in the United States in 2016, aggregated at the state level. The map was published by the *[Centers for Disease Control and Prevention](https://www.cdc.gov/)*, with data from state cancer registries.

### Map of Cancer Mortality Rates in Spain

Figure \ref{fig:choropleth-grid-create}f shows side by side maps of relative risk of lung cancer for men vs women for 2004 to 2008 [@cancerSpain].

### Atlas of Childhood Cancer in Ontario

Figure \ref{fig:choropleth-grid-create}g  displays the incidence rate of childhood cancers per 100,000 (by census division) for children aged 0-14, in Ontario from 1995 to 2004 [@OntarioPediatric].

## Additional considerations
<!-- 479 words -->
Maps cannot tell the entire story. Supplementary graphs and plots are often included to add more depth and information. Additional materials such as tables, graphs, and text explanations, support understanding and inference derived from maps, ensuring the message communicated will be consistent across a range of viewers [@CPISACA]. The many displays of statistical summaries, including dot plots, bar plots, box plots, cumulative distribution plots, scatter plots, normal probability plots, can provide alternative views of the cancer statistics. These can also display supporting statistics such as error, confidence intervals, distributions, sample or population sizes, standard deviation. When presenting cancer maps, @MACM believes the intuition derived from maps must be 'validated by rigorous statistical analyses', and the display of the supplementary statistics helps in this regard.

@@

It is extremely likely that each population area will have a different number of people. The distribution of the population residing in all areas may also be communicated in a table or histogram display [@NICR_1].
Atlases can connect the population to the land available to them by communicating population density. Alternatives to choropleth to address this are described in Section \ref{ch:alternatives} .

### Statistical uncertainty

With any number comes some uncertainty about its value. This may occur for a number of reasons: numbers are drawn from samples, errors occur in the disease reporting, values are simulated from a model to maintain privacy. Close to half of the atlases identified (42%, n=14) included some measure of uncertainty. Uncertainty is often recorded as sample or population size, standard deviation, confidence intervals, statistical significance, quantiles, credible intervals. The most common measure used to represent uncertainty were credible or confidence intervals (CIs).

Displaying the uncertainty associated with reported statistics is a vital feature of a cancer map, but it can be difficult to display effectively. The map focuses on displaying the statistic, and lacks additional space to represent the uncertainty. Providing an adjacent map or overlaying maps with symbols [@VSSDCUC] are two approaches.

### Demographics

Demographics include information regarding the age and sex distribution of the areas presented. Digital atlases allow for users to control the display, to select subsets such as males, females or those aged over 65.

### Socio-economic indicators

Socio-economic indicators, such asunemployment rates, poverty rates, remoteness, and education levels,  can explain how the experience of cancer prevalence varies for various members of a society. Few atlases provide this level of detail. Data is available from the United Nations Human Development Index and financial indicators are available from the World Bank [@IARC_3].

@@

## Cartograms
<!-- 1496 words inc table-->

<!-- What is a cartogram? -->
Choropleths imply uniformity of data across the geographic space however population densities are extremely unlikely to be uniform [@BCM]. The cartogram was developed to address this, by transforming the map to better indicate population [@ACCAC]. The resulting display more accurately communicates the impact of the disease across the population, as recorded by the statistic, at the sacrifice of geographic accuracy.

<!-- There are many alternative views of the world-->
<!-- Tobler map projections - I think this is not relevant here
Choropleths may be considered true topological maps, however, if the land mass displayed covers enough of the globe, there must be a transformation or distortion to display the land in 2D [@HTLWM]. The amount of distortion is related to the distance covered by the landmass displayed @GAMP.
World map projections reflect the frequent perspectives used to view the earth. Choropleth maps will always be distorted if they cover enough of the globe, just like photographs of the globe from space. Choropleth creation requires choosing a map projection that shows a favorable distortion of the geography for presenting the set of spatial information. Selecting a display can prevent misinterpretation of global statistics, as global maps face the challenge of equitable displays of land mass on maps [@RSCW].
If the statistic presented on the map base relies on physical distance and is influenced by the topology there is no transformation needed, beyond choosing a reasonable projection. 
-->

```{r choroCRS, fig.cap = "Four choropleth maps average age-adjusted rate of incidence for lung and bronchus in the United States averaged over sex and years 2012-2016, using four different map projections. The map projections alter the shapes and angles of the boundaries of each state.", message=FALSE, warning=FALSE, out.width='100%', fig.align='center'}
#knitr::include_graphics("figures/ggchoroCRS.png")
c1 <- png::readPNG("figures/gg3857.png")
c2 <- png::readPNG("figures/gg2163.png")
c3 <- png::readPNG("figures/gg4326.png")
c4 <- png::readPNG("figures/gg2955.png")
#knitr::include_graphics("figures/usa_grid.png")
load("figures/usa_legend.rda")

@@

<!-- Denominators -->
The purposeful distortion of the map space, transformed according to population density, is beneficial when a uniform density of the map base is desired. Population then becomes a uniformly distributed background for the statistic presented [@GOINO]. "population distribution is often extremely uneven in former British colonies" [@ACTUC ], and this makes the distortion necessary [@CTTMB].
When implementing a distortion of the geographical shape according to population, the resulting display is an area cartogram [@NAC], or population-by-area cartogram [@TAAM]. Event cartograms change the area of regions on a map depending on the amount of disease related events, but this does not consider the effects of land area and population [@VSSDCUC].

<!-- Why transform?-->
<!-- Common variables used to create cartograms, e.g. population, mortality -->
Cartograms provide an alternative visualization method for statistical and geographical information.
The key difference between a choropleth and a cartogram is the desirable augmentation of the size, shape or distance of geographical areas [@ACTUC]. 
Monmonier [@HTLWM] suggests that white lies may be employed to create useful displays and map creators have the ability to draw lines that may distort the geometry and suppress features and it is easy for the average person to disregard the impact of transformations used to create cartograms.
Cartograms may be seen as an extension of map transformations and projections. The favorable distortion is proportional to a value other other than actual earth size area [@NAC].
A disadvantage of the conventional map is that sparsely populated rural areas may be emphasized, whereas the areas representing cities are very small, making interpretation of spatial patterns very difficult.
The distortion of a cartogram accounts for the population density, preventing it from obscuring the spatial patterns [@TAAM]. 
The spatial transformation of map regions relative to the data emphasizes the data distribution instead of land size [@CBATCC].
When visualizing population statistics, Dorling [@ACTUC] considers this equitable representation design 'more socially just', or honest [@NISCC], giving due attention to all members of the population and reducing the visual impact of large areas with small populations [@DMAHP]. 
Howe [@HEDP] suggests that 'cancer occurs in people, not in geographical areas' and @CTTMB believe that spatial socio-economic data, like cancer rates, are best presented on a cartogram for urban areas as the population map base avoids allocating 'undue prominence' to rural areas. 
<!-- @MTMSIH encourage the use of cartograms to highlight small areas and uncover local-level inequalities. DI SAYS I think that focuses on bayesian analysis not on cartograms-->
